---
title: "PICtR: Physically Interacting Cell toolkit for R"
author: "Viktoria Flore"
date: "last updated 01/2026"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{PICtR: Physically Interacting Cell toolkit for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE
)
```

```{r}
library(PICtR)
```

# Introduction

Many fundamental processes in life are shaped by physical interactions between cells. PICtR provides a computational workflow to analyse both cellular landscapes and cellular interactions in high-dimensional flow or mass cytometry data using R. PICtR can be used with any cytometry-based assay and the approach can also be applied to existing datasets, provided that the following guidelines are met:  

- The data stems from a case-control setting
- All samples were treated the same during sample processing and acquisition, e.g. regarding equal cell concentrations, incubation times, and flow rates 
- Pooling across conditions, organs or donors should be avoided
- For *in vivo* experiments, cells with a strong affinity to interact can artificially interact if they are brought into proximity during sample preparation, even if they were physically separated *in vivo*. These interactions may still reflect biological effects, but this should be taken into account when interpreting the results.  

# Installation

PICtR requires R version 4.3 or later. First, please install [BPCells](https://github.com/bnprks/BPCells). Next, PICtR can be installed using:

```{r, eval=FALSE}
if (!require("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("agSHaas/PICtR")
```


Some other packages are not necessarily required, but expand the functionality of PICtR. Consider installing:  

- [flowMeans](https://www.bioconductor.org/packages/release/bioc/html/flowMeans.html)  
- [dbscan](https://github.com/mhahsler/dbscan)  
- [Spectre](https://github.com/ImmuneDynamics/Spectre)
- [ComplexHeatmap](https://www.bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)  
- [cytoMEM](https://www.bioconductor.org/packages/release/bioc/html/cytoMEM.html)  
- [autothresholdr](https://github.com/rorynolan/autothresholdr)
  
  
# PICtR Pipeline

Please refer to the flow cytometry or mass cytometry vignette for instructions. 

# Citing PICtR 

If you are using PICtR in your research, please consider citing:  

- Vonficht, Jopp-Saile, Yousefian, Flore et al. Ultra-high-scale cytometry-based cellular interaction mapping. _Nature Methods_ (2025). <https://doi.org/10.1038/s41592-025-02744-w>

# References  
  
- Vonficht, Jopp-Saile, Yousefian, Flore et al. Ultra-high-scale cytometry-based cellular interaction mapping. _Nature Methods_ (2025). <https://doi.org/10.1038/s41592-025-02744-w>
- Hao Y et al. Dictionary learning for integrative, multimodal and scalable single-cell analysis. _Nat Biotechnol_ 42, 293–304 (2024). <https://doi.org/10.1038/s41587-023-01767-y>  
- Heumos L, Schaar A.C. et al. Best practices for single-cell analysis across modalities. _Nat Rev Genet_ (2023). <https://doi.org/10.1038/s41576-023-00586-w>  
- Parks B & Abdi I. BPCells: Single Cell Counts Matrices to PCA. R package version 0.2.0 (2024) <https://bnprks.github.io/BPCells> 
