---
title: "PICtR for Mass Cytometry"
author: "Viktoria Flore"
date: "last updated 01/2026"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{PICtR for Mass Cytometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE
)
```

```{r}
library(PICtR)
```

# PICtR Pipeline

## Pre-Processing of Mass Cytometry Data 

Data from FCS files should be transformed (for example using the Arcsinh transformation of FlowJo from BD) and debris should be removed. Export the populations of interest using channel values defined by the inbuilt export function of FlowJo or use [flowWorkspace](https://github.com/RGLab/flowWorkspace). PICtR requires the DNA probe parameter(s) to be exported.

In addition to all relevant antibody-targeted parameters, we recommend using the event length parameter and all available DNA probe parameters as features during the workflow  


## Example Data

tbd

## Subsampling, Dimensional Reduction and Clustering 

tbd

### Guidelines for Clustering 

tbd

## Predict Labels for the full Data Set

tbd

## Selecting Interacting Populations

tbd

## Singlet Annotation 

tbd

## Annotation of Interacting Populations 
  
tbd

## Downstream Analysis

After annotating the singlet and interacting landscapes, you can continue with downstream analyses according to your experimental setup. Often, this will involve the quantification of interactions in the case and control conditions to test for significant differences.  
  

# Citing PICtR 

If you are using PICtR in your research, please consider citing:  

- Vonficht, Jopp-Saile, Yousefian, Flore et al. Ultra-high-scale cytometry-based cellular interaction mapping. _Nature Methods_ (2025). <https://doi.org/10.1038/s41592-025-02744-w>

# References  
  
- Vonficht, Jopp-Saile, Yousefian, Flore et al. Ultra-high-scale cytometry-based cellular interaction mapping. _Nature Methods_ (2025). <https://doi.org/10.1038/s41592-025-02744-w>
- Hao Y et al. Dictionary learning for integrative, multimodal and scalable single-cell analysis. _Nat Biotechnol_ 42, 293–304 (2024). <https://doi.org/10.1038/s41587-023-01767-y>  
- Heumos L, Schaar A.C. et al. Best practices for single-cell analysis across modalities. _Nat Rev Genet_ (2023). <https://doi.org/10.1038/s41576-023-00586-w>  
- Parks B & Abdi I. BPCells: Single Cell Counts Matrices to PCA. R package version 0.2.0 (2024) <https://bnprks.github.io/BPCells> 

